Many-core applications to online track reconstruction in HEP experiments

### 2013-10-21 papers/2013/CHEP/paper.tex
Modified abstract and intro.

\begin{abstract}
  In this paper we describe updates to a previous study of latencies
  in GPU for potential trigger applications, where we measured the
  latency to transfer data to/from the GPU, exploring the timing of
  different I/O technologies on different GPU models.  Those studies,
  where a simplified track fitting algorithm was parallelized and run
  on a GPU, show that latencies of few tens of microseconds can be
  achieved to transfer and process packets of 4 kB of data, combining
  the modern Infiniband data transfer technology with direct access to
  GPU memory allowed by NVIDIA GPUDirect utilities.  We now have
  expanded our latency studies to include other multi-core systems
  (Intel Xeon Phi/MIC in addition to NVIDIA GPUs). We also discuss the
  implementation of a scaled-up version of the algorithm used at CDF
  for online track reconstruction - the SVT algorithm - as a realistic
  test-case for low-latency trigger systems using new computing
  architectures for LHC experiment and examine the
  complexity/performance trade-off in porting existing serial
  algorithms to many-core devices.
\end{abstract}

\section{Introduction}
%%% Verbatim from NSS paper
Real-time event reconstruction plays a fundamental role in High Energy
Physics (HEP) experiments at hadron colliders.  Reducing the rate of
data to be saved on tape from millions to hundreds of events per
second is critical. To increase the purity of the collected samples,
rate reduction has to be coupled with an initial selection of the most
interesting events.  In a typical hadron collider experiment, the
event rate has to be reduced from tens of MHz to a few kHz.  The
selection system (trigger) is usually organized in multiple levels,
each capable of performing a finer selection on more complex physics
objects describing the event. Trigger systems usually comprise a first
level based on custom hardware, followed by one or two levels usually
based on farms of general purpose processors.  The possibility of also
using commercial devices at a low trigger level is very appealing:
they are subject to continuous performance improvements driven by the
consumer market, are less expensive than dedicated hardware, and are
easier to support.  Among the commercial devices, Graphic Processing
Units (GPUs) are of particular interest for online selections given
their impressive computing power (the latest NVIDIA~\cite{bib_nvidia}
GPU architecture, Kepler, exceeds Teraflop computing power); moreover,
high-level programming architectures based on C/C++ such as
\textsc{cuda}~\cite{bib_cuda} and \textsc{opencl}~\cite{bib_opencl}
make these devices more accessible to the general user.  The ultimate
goal of this study is to investigate the strengths and weaknesses of
GPUs when applied in a low latency environment, with particular
emphasis on the data transfer latency to/from the GPU and the
algorithm latency for processing on the GPU in a manner similar to a
typical HEP trigger application, and to understand the cost/complexity
ratio of porting legacy serial code to many-core devices..

We showed initial studies on GPU performance in low-latency
environments ($\sim 100~\mu$s) in a previous
paper~\cite{TIPP2011,NSS2012}.  In this paper we extend those studies
measuring the timing performance of different GPU cards in different
data I/O environments. The algorithm run on the GPU is a complete
version of the fast track fitting algorithm of the Silicon Vertex
Tracker (SVT) system at CDF~\cite{SVT1}.
%We also compare the performance of GPU
%cards from different manufacturers (AMD vs NVIDIA) and different
%computing environments (\textsc{cuda} and \textsc{opencl}) on the same
%card. 
One of the goals of the studies is to explore using a concrete
algorithm what gains one can achieve with varying degrees of changes
to the existing algorithm. Starting with a serial algorithm
implemented on a CPU, we test an \textit{embarrassingly parallel}
algorithm on the Intel MIC environment. In this case each event is
handled independently by a core on the accelerator, and the
parallelization is achieved with only minor changes to the legacy code
and is only possible in the Intel MIC environment. Next we consider an
algorithm where we unroll three internal nested loops and run these in
parallel on a GPU, using the \textsc{cuda} environment. This second
approach is programmatically more complicated and less trivial to
implement. In neither case have we re-thought the basic algorithms or
the data structures used. To achieve optimal performance, these steps
would have to be taken.  As one might expect, the improvement from the
first approach is rather modest, and the second approach shows larger
performance gains.

@@

\begin{table}[!t]
  \centering
  \begin{tabular}{|l|c|c|c|c|}
    \hline
    Model & Tesla M2050 & Tesla K20m & GeForce GTX  Titan & MIC 5110P \\
    \hline
    \hline
    Performance (SP, GFlops) & 1030 & 3520 & 4500 & 2022 \\
    Memory bandwidth  (GB/s) & 148 & 208 & 288  & 320\\   
    Memory size (GB) & 3 & 5 & 6 & 8 \\
    Number of cores & 448 & 2496 & 2688 & 240 \\
    Clock speed (GHz) & 1.15 & 0.706 & 0.837 & 1.053 \\
    \hline
  \end{tabular}
  \caption{Capabilities of the many-core devices used in this study,
    according to the manufacturer's specifications. The first three
    are GPUs and the final is an Intel Xeon Phi.}
  \label{tab_hwspecs}
\end{table}

@@

 \begin{figure}[!tbp]
   \centering
   \includegraphics[width=0.9\linewidth]{figures/Speedup_MIC}
   \caption{Speed-up with respect to a standard CPU (Intel Xeon
     E5630). The speed-ups plateau after about one million fits for
     the GPUs (in red) while the Intel MIC still shows a small slope
     beyond that point. \fixme{is this true}}
   \label{fig:algo_only_speedup}
 \end{figure}

\subsubsection{Breakdown of computing time}
In Fig.~\ref{fig:breakdown} we show the fractional time spent in
various parts of the algorithm for the embarrassingly parallel algorithm 
(on Intel MIC) and the parallel algorithm (on NVIDIA Titan GPU), as a 
function of the number of fits. On both accelerator cards the fractional times
are constant over 500 input events, where computing resources are saturated. 
Unlike the MIC, the fit stage takes most of the time on the GPU: this 
could be caused by the intense memory access frequency intrinsic to
this part of the algorithm.  
\begin{figure*}[tbp]
\centering
\subfigure[]
{\label{fig:breakdown_MIC}
\includegraphics[width=0.45\linewidth]{figures/Mic5_perc_nk}}
\subfigure[]
{\label{fig:breakdown_GPU}
\includegraphics[width=0.45\linewidth]{figures/titan_perc}}
\caption{Breakdown of computing time for MIC (a) and the GTX Titan GPU
  (b). Red corresponds to unpacking, green corresponds to generating
  hit combinations, blue is the linearized track fit, and magenta
  corresponds to offloading (MIC only)\fixme{we need a legend for
    these plots, and it doesn't work in B\&W. Also can we make them
    the same size and aspect ratio.} For MIC, combinations and fitting
  take the same amount of time for large number of events. For GPU,
  fitting dominates. }
\label{fig:breakdown}
\end{figure*}

@@

In Fig.~\ref{fig:xferlatency} we show the total latency (data
transfer, copy to/from the GPU and data processing on the GPU) as a
function of data packet size when data are transferred using GPUDirect
v1.0, \textsc{cuda}-aware MPI and P2P. For the packet sizes over 600
kB \textsc{cuda}-aware MPI gives the best performance. For smaller
packet sizes, P2P gives better performance \fixme{do we have this plot?}.

>>>>

\begin{abstract}
  Interest in parallel architectures applied to real time selections is 
  growing in High Energy Physics (HEP) experiments. In this paper we describe
  performance measurements of Graphic Processing Units (GPUs) and Intel Many
  Integrated Core architecture (MIC) when applied to a typical HEP online task: 
  the selection of events based on the trajectories of charged particles.
  We use as benchmark a scaled-up version of the algorithm used at CDF 
  experiment at Tevatron for online track reconstruction - the SVT algorithm - 
as a realistic  test-case for low-latency trigger systems using new computing
  architectures for LHC experiment. We examine the
  complexity/performance trade-off in porting existing serial
  algorithms to many-core devices. Measurements of both data processing 
  and data transfer latency are shown, considering different I/O strategies 
  to/from the parallel devices. 
\end{abstract}

\section{Introduction}
%%% Verbatim from NSS paper
Real-time event reconstruction plays a fundamental role in High Energy
Physics (HEP) experiments at hadron colliders.  Reducing the rate of
data to be saved on tape from millions to hundreds of events per
second is critical. To increase the purity of the collected samples,
rate reduction has to be coupled with an initial selection of the most
interesting events.  In a typical hadron collider experiment, the
event rate has to be reduced from tens of MHz to a few kHz.  The
selection system (trigger) is usually organized in multiple levels,
each capable of performing a finer selection on more complex physics
objects describing the event. Trigger systems usually comprise a first
level based on custom hardware, followed by one or two levels usually
based on farms of general purpose processors.  The possibility of also
using commercial devices at a low trigger level is very appealing:
they are subject to continuous performance improvements driven by the
consumer market, are less expensive than dedicated hardware, and are
easier to support.  Among the commercial devices, many-core architectures 
such as Graphic Processing Units (GPUs)~\cite{bib_gpu} and Intel Many
  Integrated Core (MIC)~\cite{bib_intelMIC} are of particular interest for 
online 
selections given their great computing power (the latest 
\textsc{nvidia}~\cite{bib_nvidia} GPU architecture, Kepler, exceeds Teraflop 
computing power); moreover,
high-level programming architectures based on C/C++ such as
\textsc{cuda}~\cite{bib_cuda}  and 
\textsc{opencl}~\cite{bib_opencl} make these devices more accessible to 
the general user.  The ultimate
goal of this study is to investigate the strengths and weaknesses of
many-core devices when applied in a low latency environment, with particular
emphasis on the data transfer latency to/from the device and the
algorithm latency for processing on the device in a manner similar to a
typical HEP trigger application, and to understand the cost/complexity
ratio of porting legacy serial code to many-core devices.

We showed initial studies on GPU performance in low-latency
environments ($\sim 100~\mu$s) in  previous
papers~\cite{TIPP2011,NSS2012}.  In this paper we extend those studies
to include other many-core systems (Intel MIC in addition to GPUs).
The algorithm run on the parallel architecture is a complete
version of the fast track fitting algorithm of the Silicon Vertex
Tracker (SVT) system at CDF~\cite{SVT1}.
%We also compare the performance of GPU
%cards from different manufacturers (AMD vs NVIDIA) and different
%computing environments (\textsc{cuda} and \textsc{opencl}) on the same
%card. 
Starting with a serial algorithm
implemented on a CPU, we test an \textit{embarrassingly parallel}
algorithm on the Intel MIC environment. In this case each event is
handled independently by a core on the accelerator, and the
parallelization is achieved with only minor changes to the legacy code
and is only possible in the Intel MIC environment. Next we consider an
algorithm where we unroll three internal nested loops and run these in
parallel on a GPU, using the \textsc{cuda} environment. This second
approach is programmatically more complicated and less trivial to
implement. In neither case have we re-thought the basic algorithms or
the data structures used. To achieve optimal performance, these steps
would have to be taken.  As one might expect, the improvement from the
first approach is rather modest, and the second approach shows larger
performance gains. For GPUs, we also test different strategies to transfer
 data to/from the device.

@@

\begin{table}[!t]
  \centering
  \begin{tabular}{|l|c|c|c|c|}
    \hline
    Model & Tesla M2050 & Tesla K20m & GeForce GTX  Titan & MIC 5110P \\
    \hline
    \hline
    Performance (SP, GFlops) & 1030 & 3520 & 4500 & 2022 \\
    Memory bandwidth  (GB/s) & 148 & 208 & 288  & 320\\   
    Memory size (GB) & 3 & 5 & 6 & 8 \\
    Number of cores & 448 & 2496 & 2688 & 240 \\
    Clock speed (GHz) & 1.15 & 0.706 & 0.837 & 1.053 \\
    \hline
  \end{tabular}
  \caption{Capabilities of the many-core devices used in this study,
    according to the manufacturer's specifications. The first three
    are \textsc{nvidia} GPUs and the final is an Intel Xeon Phi.}
  \label{tab_hwspecs}
\end{table}

@@

 \begin{figure}[!tbp]
   \centering
   \includegraphics[width=0.9\linewidth]{figures/Speedup_MIC}
   \caption{Speed-up with respect to a standard CPU (Intel Xeon
     E5630). The speed-ups plateau after about two million fits.
     %for
     %the GPUs (in red) while the Intel MIC still shows a small slope
     %beyond that point. 
     \fixme{Silvia: removed the comment about the different GPU/MI beahviour. These are the speedup values for the MIC: 0.07, 0.49, 1.79, 2.25, 2.32, 3.35, 4.38,4.74, 5.36, 5.16, 5.34 }}
   \label{fig:algo_only_speedup}
 \end{figure}

\subsubsection{Breakdown of computing time}
In Fig.~\ref{fig:breakdown} we show the fractional time spent in
various parts of the algorithm for the embarrassingly parallel algorithm 
(on Intel MIC) and the parallel algorithm (on NVIDIA Titan GPU), as a 
function of the number of fits. On both accelerator cards the fractional times
are constant over 500 input events, where computing resources are saturated. 
Unlike the MIC, the fit stage takes most of the time on the GPU: this 
could be caused by the intense memory access frequency intrinsic to
this part of the algorithm.  
\begin{figure*}[tbp]
\centering
\subfigure[]
{\label{fig:breakdown_MIC}
\includegraphics[width=0.45\linewidth]{figures/Mic5_perc_nk}}
\subfigure[]
{\label{fig:breakdown_GPU}
\includegraphics[width=0.45\linewidth]{figures/titan_perc}}
\caption{Breakdown of computing time for MIC (a) and the GTX Titan GPU
  (b). Red corresponds to unpacking, green corresponds to generating
  hit combinations, blue is the linearized track fit, and magenta
  corresponds to offloading (MIC only). For MIC, combinations and fitting
  take the same amount of time for large number of events. For GPU,
  fitting dominates. }
\label{fig:breakdown}
\end{figure*}

@@

In Fig.~\ref{fig:xferlatency} we show the total latency (data
transfer, copy to/from the GPU and data processing on the GPU) as a
function of data packet size when data are transferred using GPUDirect
v1.0, \textsc{cuda}-aware MPI and P2P. For the packet sizes over 600
kB \textsc{cuda}-aware MPI gives the best performance (see also~\cite{NSS2012} and~\cite{bib_mvapich})  \fixme{do we have this plot? Silvia: no, and will not have in time for the proceedings... but we can put a reference, see slides after page 47.}.
